Probabilistic Symbolic Execution

### 2021-06-30 shared/Paper-Draft/symbolic_execution.tex
WIP : Algorithm Section (B)

		\subsection{Background}
\label{sec:trad_sym_exec}

@@

% Symbolic Execution Algorithm.
\begin{algorithm}
	\caption{Probabilistic Symbolic Execution Algorithm}
	\label{alg:symbex}
	\begin{algorithmic}[1]
		\Function{SymbEx}{$Prog : Program$}
		\Let{$\varphi_{paths}$}{[]}, {$Ex_{stack}$} $\leftarrow$ {[]}, $\Delta \leftarrow$ {[]}, $\phi_{paths} \leftarrow \phi$, $P \leftarrow \phi$  \algorithmiccomment{Initialization}
		\Let{$I_{0}$}{\textsc{getStartInstruction}($Prog$)}
		\Let{$S_{0}$}{[$I_{0}$, $\phi_{path}$, $\Delta$, $\phi$]} \algorithmiccomment{Empty Initial State}
		\State {$Ex_{stack}$.\textsc{Push}({$S_{0}$})} \algorithmiccomment{Start with $S_{cur}$ in Execution Stack}
		\While{$Ex_{stack}$ != $\phi$}
		\Let{$S_{cur}$}{$Ex_{stack}$.\textsc{Pop}( )}, $I_{cur} \leftarrow S_{cur}[1]$ \algorithmiccomment{Start State, [$I_{0}$, $\phi_{path}$, $\Delta$]}
		\Switch{\textsc{instType}($I_{cur}$)} \label{line:inst_case}
		\Case{\texttt{make\_pse\_sym}} ($x_{var},  d_{dist}$) \algorithmiccomment{Create \texttt{fresh} Symbolic Variable} \label{line:make_pse}
		\Let{$I_{cur}$}{\textsc{getNextInstruction}($I_{cur}$)}
		\Let{$\phi_{path}, \Delta, P$}{{\textbf{PSESample}}({$x_{var},  d_{dist}$, $\phi_{path}$, $\Delta$, $P$})}
		\Let{$S_{cur}$}{[$I_{cur}$, $\phi_{path}$, $\Delta$, $P$]}
		\State {$Ex_{stack}$.\textsc{Push}({$S_{cur}$})}
		\EndCase
		\Case{v := e} \algorithmiccomment{Assignment Instruction} \label{line:assign}
		\Let{$I_{cur}$}{\textsc{getNextInstruction}($I_{cur}$)}
		\Let{$\phi_{path}, \Delta, P$}{{\textbf{PSEAssignment}}({v, e, $\phi_{path}$, $\Delta$, $P$})}
		\Let{$S_{cur}$}{[$I_{cur}$, $\phi_{path}$, $\Delta$, $P$]}
		\State {$Ex_{stack}$.\textsc{Push}({$S_{cur}$})}
		\EndCase
		\Case{\textbf{if} $(c_{sym})$ \textbf{then} $P_{1}$ \textbf{else} $P_{2}$} \algorithmiccomment{Branch Instruction}
		\Let{$Branch_{expr}$}{{\textbf{PSEBranch}}({$c_{sym}$, $\phi_{path}$, $\Delta$, $P$})} \label{line:branch}
		\Let{$I_{1}$}{\textsc{getStartInstruction}($P_{1}$)} \Let{$I_{2}$} {\textsc{getStartInstruction}($P_{2}$)}	
		\Let{$S_{true}$}{[$I_{1}$, $\phi_{path} \wedge c_{sym}$, $\Delta$, $Branch_{expr}[0]$]} \label{line:true_state}
		\Let{$S_{false}$}{[$I_{2}$, $\phi_{path} \wedge (\neg c_{sym})$, $\Delta$, $Branch_{expr}[1]$]} \label{line:false_state}
		\State {$Ex_{stack}$.\textsc{Push}({$S_{false}$})}
		\State {$Ex_{stack}$.\textsc{Push}({$S_{true}$})} \algorithmiccomment{Start with True State}
		\EndCase
		\Case{\textbf{HALT}} \algorithmiccomment{Terminate Instruction}
		\Let{$\varphi_{paths}$}{$\varphi_{paths}$.\textsc{Append}($\phi_{path}$)}		
		\EndCase
		\EndSwitch
		\EndWhile \\
		\Return{$\varphi_{paths}$}
		\EndFunction
	\end{algorithmic}
\end{algorithm}

\textit{Symbolic execution} is a program analysis technique where a program is run on \textit{symbolic} inputs and all program operations are replaced with those which manipulate these symbolic variables.
%
We present a high-level description of how \textit{symbolic execution} works in Alg~\ref{alg:symbex}. 
% 
For each path in the program, a list of \textit{constraints} are stored that encodes the whole path as a \textit{logical}
formula.
% 
These constraints are over \textit{program} variables that have been marked \textit{symbolic} by the user and are stored in the \textit{state} data structure as \textit{metadata} during the \textit{dynamic} symbolic execution of the program.
% 
Apart from \textit{path constraints}, the \textit{state} also contains a mapping of \textit{symbolic} variables to expressions that are either  \textit{symbolic} or \textit{concrete} (in the case of concrete execution of the program) and a list of \textit{instructions} currently getting executed as a part of the \textit{state}.	
% 
During symbolic execution, program state is encoded symbolically in two parts: a conjunctive formula, $\varphi$, consisting of the branch conditions which are true for that particular path, called a \textit{path condition}, and a mapping from program variables to symbolic expressions containing constants and \textit{symbolic variables}, $\sigma$.
% 
Each \textit{instruction} in the program is executed and depending upon the \textit{type} of the \textit{instruction} the \textit{state} along with the \textit{symbolic variables} mapping, ($\Delta$ in Alg~\ref{alg:symbex}) are updated as described in Algorithm~\ref{alg:symbex}.
% 
\textit{Symbolic variables} are assigned values by employing an \textsc{SMT Solver} for solving the current set of \textit{constraints} collected so far in the \textit{program path} and also by \textit{concretizing} some of the values that get evaluated as a result of \textit{concrete} execution (e.g. return value from external function calls).
%
When execution reaches an assignment of the form $\mathtt{x} = e$, where $e$ is a constant or program variable, $\sigma[\mathtt{x}]$ is either just the constant, $e$, or the symbolic representation of $e$, namely $\sigma[e]$.
% 
If $e$ is an expression, we recursively convert each program variable in $e$ to its corresponding symbolic expression found in $\sigma$ to construct a new symbolic execution, $e_{sym}$, and set $\sigma[\mathtt{x}] = e_{sym}$.
% 
When execution reaches a branch guarded by the condition $c$, execution proceeds down both branches, one where $\varphi = \varphi \wedge \sigma[c]$, and the other where $\varphi = \varphi \wedge \neg \sigma[c]$.

\subsection{Adding Probabilistic Sampling}
We present the formalism of our approach on \textbf{pWhile}, a core imperative probabilistic programming language, as a model for more general languages:
% 
\[
S := \mathbf{skip} \mid \mathtt{x} \leftarrow e \mid \mathtt{x} \xleftarrow{\$} d \mid S_1 ; S_2 \mid \mathbf{if}~e~\mathbf{then}~S_1~\mathbf{else}~S_2 \mid \mathbf{while}~e~\mathbf{do}~S 
\]

@@

\begin{paragraph}{Assignment}
	For assignment statements of the form $\mathtt{x} \leftarrow e$, where $\mathtt{x}$ is a \textit{program} variable and $e$ is an expression, probabilistic symbolic execution proceeds identically to traditional symbolic execution, as detailed in Alg.~\ref{alg:assign}.
	%
	On the \textit{symbolic execution} side, the \textit{symbolic variable} mapping ($\Delta$) is updated with the result of the assignment operation is stored back in the current \textit{state} for further execution as shown in Line~\ref{line:assign} of Alg~\ref{alg:symbex}.
\end{paragraph}

@@

\begin{algorithm}
	\caption{PSE Branch Algorithm}
	\label{alg:branch}
	\begin{algorithmic}[1]
		\Function{PSEBranch}{$c, \varphi, \sigma, P$}
		\Let{$c_{sym}$}{$\sigma[c]$}
		\Let{$(\delta_1,\ldots,\delta_n)$}{$\dom{P}$}
		\Let{$(d_1,\ldots,d_n)$}{$(P[\delta_1],\ldots,P[\delta_n])$}
		\Let{$p_c$}{$\frac{\displaystyle\sum_{(v_1,\ldots,v_n) \in \dom{d_1} \times \cdots\times \dom{d_n}} [(c_{sym} \wedge \varphi)\{\delta_1 \mapsto v_1,\ldots,\delta_n \mapsto v_n\}]}{\displaystyle\sum_{(v_1,\ldots,v_n) \in \dom{d_1} \times \cdots\times \dom{d_n}}[\varphi\{\delta_1 \mapsto v_1,\ldots,\delta_n \mapsto v_n\}]}$}
		\State\Return{$((\varphi \wedge c_{sym}, \sigma,P,p_c), (\varphi \wedge \neg c_{sym}, \sigma,P,1-p_c))$}
		\EndFunction
	\end{algorithmic}
\end{algorithm}

>>>>

@@

% Symbolic Execution Algorithm.
\begin{algorithm}
	\caption{Probabilistic Symbolic Execution Algorithm}
	\label{alg:symbex}
	\begin{algorithmic}[1]
		\Function{SymbEx}{Prog : Program}
		\Let{$\varphi_{paths}$}{[]}, {$L$} $\leftarrow$ {[]}, $Enc_t \gets \bot, Enc_a \gets \bot$  \algorithmiccomment{Initialization}
		\Let{$I_{0}$}{\textsc{getStartInstruction}($Prog$)}
		\Let{$S_{0}$}{($I_{0}$, $\phi$, $\phi$, $\phi$, 1.0)} \algorithmiccomment{Empty Initial State}
		\State {$L$.\textsc{Append}({$S_{0}$})} \algorithmiccomment{Start with $S_{cur}$ in Execution Stack}
		\While{$L \neq \phi$}
		\Let{($I_{cur}, \varphi, \sigma, P, p)$}{$L$.\textsc{Remove}( )}
		\If{\textsc{UNSAT}($\varphi$) $\lor p = 0.0$}
		\State \textbf{continue}
		\EndIf
		\Switch{\textsc{instType}($I_{cur}$)} \label{line:inst_case}
		\Case{$x \gets d$} \algorithmiccomment{Create \texttt{fresh} Symbolic Variable} \label{line:make_pse}
		\Let{$I_1$}{\textsc{getNextInstruction}($I_{cur}$)}
		\Let{$P_1$}{{\textbf{PSESample}}({$x,  d, P$})}
		\Let{$S_1$}{($I_{cur}, \varphi, \sigma, P_1, p$)}
		\State {$L$.\textsc{Append}({$S_1$})}
		\EndCase
		\Case{$v \gets e$} \algorithmiccomment{Assignment Instruction} \label{line:assign}
		\Let{$I_1$}{\textsc{getNextInstruction}($I_{cur}$)}
		%\Let{$e_{sym}$}{$\sigma[e]$}
		%\Let{$\varphi, \sigma, P$}{{\textbf{PSEAssignment}}({v, e, $\varphi$, $\sigma$, $P$})}
		\Let{$\sigma[x]$}{$\sigma[e]$}
		\Let{$S_{cur}$}{($I_{cur}$, $\varphi$, $\sigma$, $P$)}
		\State {$L$.\textsc{Append}({$S_{cur}$})}
		\EndCase
		\Case{\textbf{if} $(c)$ \textbf{then goto} $T_{1}$} \algorithmiccomment{Branch Instruction}
		\Let{$c_{sym}$}{$\sigma[c]$}
		\Let{$p_t, p_f$}{{\textbf{PSEBranch}}({$c_{sym}$, $\sigma$, $P$})} \label{line:branch}
		\Let{$I_{1}$}{\textsc{getInstruction}($T_{1}$)} 
		\Let{$I_{2}$} {\textsc{getNextInstruction}(I)}	
		\Let{$S_{true}$}{($I_{1}$, $\varphi \wedge c_{sym}$, $\sigma$, $p_t$)} \label{line:true_state}
		\Let{$S_{false}$}{($I_{2}$, $\varphi \wedge \neg c_{sym}$, $\sigma$, $p_f$)} \label{line:false_state}
		\State {$L$.\textsc{Append}({$S_{false}$})}
		\State {$L$.\textsc{Append}({$S_{true}$})} \algorithmiccomment{Start with True State}
		\EndCase
		\Case{\textbf{halt(c)}} \algorithmiccomment{Terminate Instruction}
		\Let{$c_{sym}$}{$\sigma[c]$}
		\If{\textsc{SAT}($c_{sym}$)}
		\State {$Enc_t$.\textsc{AddEncoding}($\varphi, p$)}		
		\EndIf
		\State {$Enc_a$.\textsc{AddEncoding}($\varphi, p$)}		
		\EndCase
		\EndSwitch
		\EndWhile 
		\Let{$r$}{\textsc{Solve}($Enc_t, Enc_a$)} \\
		\Return{$r$}
		\EndFunction
	\end{algorithmic}
\end{algorithm}

\textit{Symbolic execution} is a program analysis technique where a program is run on \textit{symbolic} inputs and all program operations are replaced with those which manipulate these symbolic variables.
%
We present a high-level description of how \textit{symbolic execution} works in Alg~\ref{alg:symbex}. 
% 
For each path in the program, a list of \textit{constraints} are stored that encodes the whole path as a \textit{logical}
formula.
% 
These constraints are over \textit{program} variables that have been marked \textit{symbolic} by the user and are stored in the \textit{state} data structure as \textit{metadata} during the \textit{dynamic} symbolic execution of the program.
% 
Apart from \textit{path constraints}, the \textit{state} also contains a mapping of \textit{symbolic} variables to expressions that are either  \textit{symbolic} or \textit{concrete} (in the case of concrete execution of the program) and a list of \textit{instructions} currently getting executed as a part of the \textit{state}.	
% 
During symbolic execution, program state is encoded symbolically in two parts: a conjunctive formula, $\varphi$, consisting of the branch conditions which are true for that particular path, called a \textit{path condition}, and a mapping from program variables to symbolic expressions containing constants and \textit{symbolic variables}, $\sigma$.
% 
Each \textit{instruction} in the program is executed and depending upon the \textit{type} of the \textit{instruction} the \textit{state} along with the \textit{symbolic variables} mapping, ($\sigma$ in Alg~\ref{alg:symbex}) are updated as described in Algorithm~\ref{alg:symbex}.
% 
\textit{Symbolic variables} are assigned values by employing an \textsc{SMT Solver} for solving the current set of \textit{constraints} collected so far in the \textit{program path} and also by \textit{concretizing} some of the values that get evaluated as a result of \textit{concrete} execution (e.g. return value from external function calls).
%
When execution reaches an assignment of the form $\mathtt{x} = e$, where $e$ is a constant or program variable, $\sigma[\mathtt{x}]$ is either just the constant, $e$, or the symbolic representation of $e$, namely $\sigma[e]$.
% 
If $e$ is an expression, we recursively convert each program variable in $e$ to its corresponding symbolic expression found in $\sigma$ to construct a new symbolic execution, $e_{sym}$, and set $\sigma[\mathtt{x}] = e_{sym}$.
% 
When execution reaches a branch guarded by the condition $c$, execution proceeds down both branches, one where $\varphi = \varphi \wedge \sigma[c]$, and the other where $\varphi = \varphi \wedge \neg \sigma[c]$.

\subsection{Adding Probabilistic Sampling}
We present the formalism of our approach on \textbf{pWhile}, a core imperative probabilistic programming language, as a model for more general languages:
% 
\[
%S := \mathbf{skip} \mid \mathtt{x} \leftarrow e \mid \mathtt{x} \xleftarrow{\$} d \mid S_1 ; S_2 \mid \mathbf{if}~e~\mathbf{then}~S_1~\mathbf{else}~S_2 \mid \mathbf{while}~e~\mathbf{do}~S 
S := \mathtt{x} \leftarrow e \mid \mathtt{x} \xleftarrow{\$} d \mid S_1 ; S_2 \mid \mathbf{if}~e~\textbf{then goto} T \mid \textbf{halt}(e)
\]

@@

\begin{paragraph}{Assignment}
	For assignment statements of the form $\mathtt{x} \leftarrow e$, where $\mathtt{x}$ is a \textit{program} variable and $e$ is an expression, probabilistic symbolic execution proceeds identically to traditional symbolic execution, as detailed in Alg.~\ref{alg:assign}.
	%
	On the \textit{symbolic execution} side, the \textit{symbolic variable} mapping ($\sigma$) is updated with the result of the assignment operation is stored back in the current \textit{state} for further execution as shown in Line~\ref{line:assign} of Alg~\ref{alg:symbex}.
\end{paragraph}

@@

\begin{algorithm}
	\caption{PSE Branch Algorithm}
	\label{alg:branch}
	\begin{algorithmic}[1]
		\Function{PSEBranch}{$c, \varphi, \sigma, P$}
		\Let{$(\delta_1,\ldots,\delta_n)$}{$\dom{P}$}
		\Let{$(d_1,\ldots,d_n)$}{$(P[\delta_1],\ldots,P[\delta_n])$}
		\Let{$p_c$}{$\frac{\displaystyle\sum_{(v_1,\ldots,v_n) \in \dom{d_1} \times \cdots\times \dom{d_n}} [(c_{sym} \wedge \varphi)\{\delta_1 \mapsto v_1,\ldots,\delta_n \mapsto v_n\}]}{\displaystyle\sum_{(v_1,\ldots,v_n) \in \dom{d_1} \times \cdots\times \dom{d_n}}[\varphi\{\delta_1 \mapsto v_1,\ldots,\delta_n \mapsto v_n\}]}$}
		\State\Return{$((\varphi \wedge c_{sym}, \sigma,P,p_c), (\varphi \wedge \neg c_{sym}, \sigma,P,1-p_c))$}
		\EndFunction
	\end{algorithmic}
\end{algorithm}
